### 2024-11-15 teaching/math240f24/draft.tex
Added link

However, subgraphs are not exactly the right notion to be considering
when talking about planarity. To see why, consider the graph in Fig.~10.
\midinsert
$$\epsfbox{k5dot.ps}$$
\vskip5pt
\caption{The graph $K_5$, with an extra vertex subdividing an edge.}
\endinsert
\goodbreak
It is nonplanar, since if it had a planar embedding, then by contracting one of the
edges incident on the extra vertex and bending other edges accordingly,
we would obtain a planar embedding of $K_5$.
On the other hand, $K_5$ is not a subgraph of this graph.
We thus introduce the following notion.
We say that a graph $H$ is a {\it graph minor} of a graph $G$ if $H$ can be obtained
from $G$ by repeatedly either
\medskip
\item{i)} deleting an edge;
\smallskip
\item{ii)} deleting a vertex; or
\smallskip
\item{iii)} contracting an edge $uv$ by removing it and merging $u$ and $v$ into a single vertex
(and also combining any resulting multiple edges into a single edge).
\medskip
Any number of the first two operations simply creates a subgraph. It is the contraction operation
that produces interesting examples of graph minors. It is easy to see pictorially that
contracting edges preserves planarity (we will not prove this rigorously), so {\sl any graph
minor of a planar graph is planar}. We can use this to show, for instance, that the Petersen graph,
depicted in Fig.~11, is nonplanar, since by contracting each of the five edges connecting
the inner star pentagram to the outer pentagon, we obtain $K_5$. (Note that
Theorem~{\thmnonplanarconditions} does not directly apply to the Petersen graph.)
\midinsert
$$\epsfbox{petersen.ps}$$
\vskip5pt
\caption{The Petersen graph.}
\endinsert
\goodbreak
\bye

@@

>>>>

However, subgraphs are not exactly the right notion to be considering
when talking about planarity. To see why, consider the graph in Fig.~10.
\midinsert
$$\epsfbox{k5dot.ps}$$
\vskip5pt
\caption{The graph $K_5$, with an extra vertex subdividing an edge.}
\endinsert
\goodbreak
It is nonplanar, since if it had a planar embedding, then by contracting one of the
edges incident on the extra vertex and bending other edges accordingly,
we would obtain a planar embedding of $K_5$.
On the other hand, $K_5$ is not a subgraph of this graph.
We thus introduce the following notion.
We say that a graph $H$ is a {\it graph minor} of a graph $G$ if $H$ can be obtained
from $G$ by repeatedly either
\medskip
\item{i)} deleting an edge;
\smallskip
\item{ii)} deleting a vertex; or
\smallskip
\item{iii)} contracting an edge $uv$ by removing it and merging $u$ and $v$ into a single vertex
(and also combining any resulting multiple edges into a single edge).
\medskip
Any number of the first two operations simply creates a subgraph. It is the contraction operation
that produces interesting examples of graph minors. It is easy to see pictorially that
contracting edges preserves planarity (we will not prove this rigorously), so {\sl any graph
minor of a planar graph is planar}. We can use this to show, for instance, that the Petersen graph,
depicted in Fig.~11, is nonplanar, since by contracting each of the five edges connecting
the inner star pentagram to the outer pentagon, we obtain $K_5$. (Note that
Theorem~{\thmnonplanarconditions} does not directly apply to the Petersen graph.)
\midinsert
$$\epsfbox{petersen.ps}$$
\vskip5pt
\caption{The Petersen graph.}
\endinsert
\goodbreak

@@

To count strings of length $5$ and containing at least one $a$, we let the universe
be all strings of length $5$ and let $A$ be the set of all strings without
any $a$, i.e., using only the letters $b$, $c$, and $d$. Then the number of strings
containing at least one $a$ is
$$|\bar A| = |U| - |A| = 4^5 - 3^5.$$

Next we count the number of strings of length $5$, but we require that adjacent symbols
be different. This one is best calculated by counting the number of choices we have at
each slot. There are $4$ choices for the first symbol, but after
that we only have choices for each letter, since they cannot match the previous one.
Hence the count is $4\cdot 3^4$.

Lastly, to count the number of strings that must have at least one of the patterns
$aa$, $bb$, $cc$, or $dd$, we simply note that this is the complement of what
we counted in the previous paragraph, so the answer is $4^5 - 4\cdot 3^4$.

\medskip\boldlabel The principle of inclusion and exclusion. As our next toy problem,
suppose we want to count the number of strings of length $6$, using the symbols
$a$, $b$, $c$, and $d$, that start with $a$ or end with $aa$ (or both). The number
of strings that start with $a$ is $4^5$, since we have no choice for the first symbol,
but $4$ choices for each of the remaining five symbols; likewise, the number
of strings that end with $aa$ is $4^4$. Adding these two up, we get $4^5+4^4$.
There's a problem though. We have accidentally double-counted the strings that
both start with $a$ {\it and} end with $aa$. How many of these are there? Well, the first, fifth, and
sixth slots all have to be $a$, but there are four choices for each of the remaining
three slots, for a total of $4^3$. Subtracting this from our subtotal yields
the correct answer: $4^5+4^4-4^3$.
\endgroup%\headline{\hfil}\footline{\hfil}

This technique works with any two sets $A$ and $B$,
i.e., $|A\cup B| = |A| + |B| - |A\cap B|$. For three sets $A$, $B$, and $C$,
we have
$$|A\cup B\cup C| = |A| + |B| + |C| - |A\cap B| - |A\cap C| - |B\cap C| + |A\cap B\cap C|$$
(draw a Venn diagram and convince yourself that this is correct).
For a general number of sets, we have the following theorem.

\proclaim Theorem {\advthm} (Principle of inclusion and exclusion).
